If the Null Fits, You Must Omit: Ubiquitous False Detections of Group Polarization

### 2025-08-14 GroupPolarizationStatmod/paper/intro-latest.tex
further cleanup, expansion, and connection of intro and methods

Paragraph tres: How this procedure has been used in the great pointless theoretical milieu
that raged for decades with apparently no strong determination on the outcome
or even what the fight was all about.

Paragraph quatro e final: our corpus.

>>>>

\subsection{How study design can induce false detections of group polarization}
\label{ssubsec:false-detections}

An opinion is never measured directly, only a behavior that indicates an opinion
in a behavioral experiment or survey.  For example, the location of a mark
that a participant makes is measured, not the complex neural activity that
underlies forming opinions.
Theoretically, an internal opinion in the form of neural activity is a
\textit{latent opinion}, from the Latin \emph{latentem} meaning ``lie
hidden''.  Participants report \text{observed opinions}, typically as one of
a finite number of ranked, successive integer values that form the
\emph{selection set}. The minimum possible value in the selection set might
mean ``strongly disagree'' and the maximum value represents ``strongly
agree''.

Group polarization experiments share a common experimental design where,
first, participant opinions are gathered on a topic of interest. Next, groups
are formed that have some initial opinion extremism. Participants deliberate
or discuss the topic for some set time. After the set time, participants
again give their opinion. Group polarization is said to occur if the
mean group opinion became more extreme, i.e., if the magnitude of the mean
group opinion is greater \emph{post-deliberation} compared to
\emph{pre-deliberation}. For example,
\cite{Moscovici1969} asked Parisian high school students to indicate their
degree of agreement with the statement, ``American economic aid is always
used for political pressure''. There
were seven options in the selection set, -3 indicating ``strongly disagree'',
+3 indicating ``strongly agree'', and 0 was neutral. Moscovici and Zavalloni
report that the mean pre-deliberation opinion was -0.61 and the mean
post-deliberation opinion was -1.09 (Moscovici and Zavalloni, 1969; Table 4,
p. 132). This is a -0.48 shift to more extreme disagreement with the 
statement, reported by Moscovici and Zavalloni as statistically significant
in Table 5 on p. 132.

(PREVIOUS TWO PARAGRAPHS SET UP THIS FINAL PARAGRAPH: CONTINUE TO USE MOSCOVICI 1969 TO
EXPLAIN IN THE SIMPLEST TERMS HOW FAILING TO ACCOUNT FOR MEASUREMENT PROCEDURE
CAN LEAD TO FALSE POSITIVES; see Figure \ref{fig:distros})

### 2025-08-14 GroupPolarizationStatmod/paper/ms.tex
further cleanup, expansion, and connection of intro and methods

To find examples of plausibly spurious group polarization we guess at what
latent mean, combined with a latent pre- and post-deliberation 
latent standard deviations, could generate it. We calculate the observed group
polarization for these latent parameters by transforming the latent pre- and
post-deliberation distributions from continuous normal distributions to ordinal
"observations" by integrating over each ordinal bin, then calculating the observed
mean for each experiment *phase*, i.e., both the pre- and post-deliberation phases. 
If the null model can generate group polarization in this case, then the detection
is labelled *plausibly false*, i.e., *plausibly spurious*, and we call the 
generating latent parameters the *plausible null parameters*.

To evaluate whether the plausible null parameters we identify could realistically
fool a significance test, we implemented a generative simulation of each plausibly
false detection of group polarization and calculated the *family-wise error rate
(FWER)* and the *false discovery rate (FDR)*. We use the published number of
participants $N$ for each of the 54 conditions that yielded plausibly false
detections of group polarization. The FWER is also known as the Type I
error, or the false positive rate: it is the probability that a detection of an
effect occurs given that in reality there was a null effect.  The FDR is the
probability that a published positive detection of an effect is a false positive,
i.e., a detection of a null effect. High FWER and FDR for an experimental condition
indicates that its design is weak, since we identify only one of perhaps infinitely
many latent parameter combinations that can generate spurious group polarization.

We use this setup to demonstrate how *adversarial* style testing of experimental
designs with counterfactual null models can be used to estimate important
metascience parameters. In this study we use our FWER calculations to estimate
the minimum Cohen's $d$ that should be used to determine whether an effect is
``significant'' or not that will limit the FWER to 5\%. In the Discussion we
sketch how our model can then be extended to rigorously calculate statistical power,
i.e., the probability that a significant effect is detected given that there 
is indeed a significant effect.

@@

We can now define spurious group polarization and give examples of when it occurs.
First, the observed group polarization opinion shift is
\begin{equation}
  s = \meanobs_\mathrm{post} - \meanobs_\mathrm{pre}~,
  \label{eq:group-polarization}
\end{equation}
\noindent
where the expected value of observed opinions at time $t$ is
\begin{equation}
  \meanobst = \frac{1}{B} \sum_{b=1}^B b \cdot p(b; \mu_t, \sigma_t, B).
  \label{eq:meanobs}
\end{equation}
\noindent
\emph{Spurious} group polarization is when mean observed opinions seem to increase in 
extremity when the latent mean is held constant. We simulate spurious
group polarization by setting $\mupre = \mupost = \mu$ and setting the variances
to values known to induce spurious group polarization. 
We explain how to find a constant latent mean, $\mu$, and pre- and post-deliberation
variances, $\sigmapre$ and $\sigmapost$, respectively, that generate spurious 
detections of group polarization in the next section. 
% We describe how we find $\mu$,
% $\sigmapre$ and $\sigmapost$ next in Section~\ref{sec:finding-variances}.

>>>>

To find examples of plausibly spurious group polarization we guess at what latent
mean, combined with a latent pre- and post-deliberation latent standard deviations,
could generate it. We calculate the observed group polarization for these latent
parameters by transforming the latent pre- and post-deliberation distributions from
continuous normal distributions to ordinal ``observations'' by integrating over each
ordinal bin, then calculating the observed mean for each experiment \emph{phase},
i.e., both the pre- and post-deliberation phases.  If the null model can generate
group polarization in this case, then the detection is labelled \emph{plausibly
false}, i.e., \emph{plausibly spurious}, and we call the generating latent
parameters the \emph{plausible null parameters}.

To evaluate whether the plausible null parameters we identify could realistically
fool a significance test, we implemented a generative simulation of each plausibly
false detection of group polarization and calculated the \emph{family-wise error rate
(FWER)} and the \emph{false discovery rate (FDR)}. We use the published number of
participants $N$ for each of the 54 conditions that yielded plausibly false
detections of group polarization. The FWER is also known as the Type I
error, or the false positive rate: it is the probability that a detection of an
effect occurs given that in reality there was a null effect.  The FDR is the
probability that a published positive detection of an effect is a false positive,
i.e., a detection of a null effect. High FWER and FDR for an experimental condition
indicates that its design is weak, since we identify only one of perhaps infinitely
many latent parameter combinations that can generate spurious group polarization.

We use this setup to demonstrate how \emph{adversarial} style testing of experimental
designs with counterfactual null models can be used to estimate important
metascience parameters. In this study we use our FWER calculations to estimate
the minimum Cohen's $d$ that should be used to determine whether an effect is
``significant'' or not that will limit the FWER to 5\%. In the Discussion we
sketch how our model can then be extended to rigorously calculate statistical power,
i.e., the probability that a significant effect is detected given that there 
is indeed a significant effect.

@@

We can now define spurious group polarization and give examples of when it occurs.
First, the observed group polarization opinion shift is
\begin{equation}
  g = \meanobs_\mathrm{post} - \meanobs_\mathrm{pre}~,
  \label{eq:group-polarization}
\end{equation}
\noindent
where the expected value of observed opinions at time $t$ is
\begin{equation}
  \meanobst = \frac{1}{B} \sum_{b=1}^B b \cdot p(b; \mu_t, \sigma_t, B).
  \label{eq:meanobs}
\end{equation}
\noindent

\emph{Spurious} group polarization is when mean observed opinions seem to increase in 
extremity when the latent mean is held constant. Call the collection of
latent null model distribution parameters $\lambda = (\mu, \sigma_{pre}, \sigma_{post})$.
The group polarization opinion shift can then be written as a function of $\lambda$,
$g(\lambda)$. Formally, then, spurious group polarization is plausible if 
there exists at least one $\lambda$ for which $g(\lambda) = g_{pub} > 0$, where
$g_{pub}$ is the published group polarization shift value.

Next, we explain how to find a set of latent null parameters, $\lambda$, that
generate spurious group polarization when naively measured on an ordinal scale.

\begin{figure}
  \caption{\textbf{Opinion measurement model and spurious group polarization.} 
  Latent model participant opinions are drawn from a normal distribution with the
  listed parameters (A,C). The pre-deliberation latent distribution has high
  variance, but appears polarized when measured via integration (A).  The latent
  distribution's variance is expected to decrease during deliberation (B) via
  consensus, but in this example the latent mean is static, 
$\mupre = \mupost$. However, when the post-deliberation distribution is measured,
the simulated observed mean increases, i.e., spurious
group polarization occurs (C).}
  \centering
  \includegraphics[width=0.95\textwidth]{Figures/Model/latent-ordinal-distros.pdf}
  \label{fig:distros}
\end{figure}
